---
title: "Using DRpower for study design"
author: "Bob Verity"
date: "Last updated: `r format(Sys.Date(), '%d %b %Y')`"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{rationale6_master_protocol}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(DRpower)
library(kableExtra)
library(dplyr)
library(ggplot2)
```

## 6. The 2020 WHO master protocol

In 2020, the WHO released the [master protocol for surveillance of pfhrp2/3 deletions and biobanking to support future research](https://tinyurl.com/73ds8a67)[-@world2020master]. This document gave guidance on a wide range of issues related to *pfhrp2/3* studies, including data collection and fieldwork, data storage and laboratory analysis, and providing templates for various forms such as informed consent forms for participants. Overall, this document was a major step forward

Contains a sample size calculation:

$n \geq z_{\tiny 1 - \alpha/2}^2 \frac{\displaystyle p(1-p)}{\displaystyle (p-\mu)^2}D_{eff}$

Text

$CI_{lower} = \hat{p} - z_{\tiny 1 - \alpha/2}\sqrt{\frac{\displaystyle \hat{p}(1-\hat{p})}{\displaystyle n}D_{eff}}$

Text

$Z = \frac{\displaystyle |\hat{p} - \mu|}{\displaystyle \sqrt{\frac{\hat{p}(1-\hat{p})}{n}D_{eff}}}$

Text

$\text{Power} = \Phi\Bigg( \frac{|\hat{p} - \mu|}{\sqrt{\frac{\hat{p}(1-\hat{p})}{n}D_{eff} }} - z_{\tiny 1 - \alpha/2} \Bigg)$

Text

$\text{Power} = \Phi(0) = 0.5$

Text

$n \geq (z_{\tiny 1 - \beta} + z_{\tiny 1 - \alpha/2})^2 \frac{\displaystyle p(1-p)}{\displaystyle (p-\mu)^2}D_{eff}$

Text

$(1 + z_{\tiny 1 - \beta}/z_{\tiny 1 - \alpha/2})^2$

which equals `r round((1 + qnorm(0.8) / qnorm(0.975))^2, 3)`.

## References


